---
title: "Exploratory Data Week 2 Project"
author: "Benjamin J. Hooker"
date: "September 25, 2017"
output: html_document
---
Load packages for downloading files.  I find the `curl` pacakge to be quite 
helpful when downloading "https" files
``` {r,librariestoload}

library(curl)
library(data.table)
```
Download the file, unzip it, and then read it with `read.csv` function to a 
variable called "data"
``` {r,downloading}  

file_url <- ("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip")  
curl_download(file_url, destfile = "./data/week2_proj.zip")  
data <- read.csv(unzip("./data/week2_proj.zip"))

```
### Building the histogram:  
A simple histogram of the data using the `hist` function:
``` {r,simplehist}
hist(data$steps, main = "Historam of Steps", xlab = "Number of Steps")

```

When exploring the data using the `table` function, there are 11K+ 0's recorded
by the device, so it may make more sense to extract the non-zero numbers using
the `which` function.
``` {r,thehistogram}  

hist(which(data$steps > 0), col = "aquamarine", xlab = "Number of Steps",main = "Histogram of non-zero steps")

```

### Calculating the mean and median:
This is probably a good time to use the `tapply` function to generate some data-
frames and then perform a `cbind` to make one nice table.  
``` {r,thetapply}

data_average <- as.data.frame(with(data, tapply(steps, date, mean)))
data_med <- as.data.frame(with(data, tapply(steps, date, median)))
data_table <- cbind(data_average, data_med)
colnames(data_table) <- c("Average", "Median")
data_table
```

### Plotting steps by interval
Here again I will use the `tapply` function to generate the time-series analysis
to try to discover the mean of the steps taken at a given time interval for each
day.  I had to use an `na.omit` function earlier than expected to display a 
functional plot.
``` {r,the plot by interval}
interval_average <- as.data.frame(with(na.omit(data), tapply(steps, interval, mean)))  
plot(interval_average, type = "l", col = 4, xlab = "Time interval (by 5 min)",ylab = "Average Steps", main = "Steps by Time interval")
```

I think I have a fairly good idea of where the highest activity interval is, but 
I will run an `order` function on the new data frame to see which interval had 
the greatest number of steps.
``` {r,intervaltable}
df <- interval_average[order(-interval_average),]
head(df)
```

### Missing Data
Calculating the number of "NAs"
``` {r,countNAs}
sum(is.na(data))
```

### Making a data set without "NAs"
Here I will use the `na.omit` and create a data set that is free of NA's.
```{r,cleandata}
data1 <- na.omit(data)
sum(is.na(data1)) ## Check to ensure there are no "NAs"
```

### Making another Histogram
With NA's removed, I will remake the previous histograms, but it might be more
useful to compare them side by side:
``` {r,histoplots}
par(mfrow = c(2,2))
hist(data$steps,main = "Historam of Steps Data", xlab = "Number of Steps")
hist(data1$steps, col = 3, main = "Historam of Steps Data1",xlab = "Number of Steps")
hist(which(data$steps > 0), col = "aquamarine", xlab = "Number of Steps",main = "Histogram of non-zero steps Data")
hist(which(data1$steps > 0), col = "blue", xlab = "Number of Steps",main = "Histogram of non-zero steps Data1")
```

### Another Table
Here I will use the `tapply` function again, but the code will use the "data1" 
set instead of the original "data" set.
``` {r,thetapplyagain}

data_average1 <- as.data.frame(with(data1, tapply(steps, date, mean)))
data_med1 <- as.data.frame(with(data1, tapply(steps, date, median)))
data_table1 <- cbind(data_average1, data_med1)
colnames(data_table1) <- c("Average1", "Median1")
data_table1 <- na.omit(data_table1)
data_table1
```

### Making Weekends
First I need to convert the information in the data set to acutal dates using 
the `as.Date` function in conjunction with `as.character` function and the 
"format = '%Y-%m-%d'" option.
``` {r,makethedate}
data1$date <- as.Date(as.character(data1$date), "%Y-%m-%d")
```

Now I need to create a factor variable in another column to create a table to 
diferentiate between the two types of days.
``` {r,forloop}
data1[,4] <- weekdays(data1$date)
for(i in 1:nrow(data1)){
        if (data1[i,4]== "Saturday" | data1[i,4]== "Sunday")
        {data1[i,5]<-"Weekend"}
        else {data1[i,5]<-"weekday"
                }
}
```

Weekend factor created, now it is time to make it a factor.
``` {r,thefactor}
data1$V5 <- as.factor(data1$V5)
names(data1)[5] <- "Weekend"
```

Now it is time to subset the data into "Weekdays" and "Weekends"
``` {r,subsetting}
weekend <- data1[data1$Weekend == "Weekend",]
weekday <- data1[data1$Weekend == "weekday",]
```

Run some `tapply` functions like earlier on the new data frames and then plot 
the results per the assignment.
``` {r,plotting again}
interval_average1 <- as.data.frame(with(na.omit(weekday), tapply(steps, interval,mean)))
interval_average2 <- as.data.frame(with(na.omit(weekend), tapply(steps, interval,mean)))
par(mfrow = c(2,1))
plot(interval_average1, type = "l", col = 2, xlab = "Intervals (5 min)",ylab = "Average number of steps", main = "Weekday steps")
plot(interval_average2, type = "l", col = 4, xlab = "Intervals (5 min)",ylab = "Average number of steps", main = "Weekend steps")
